---
title: "PA1_template"
author: "Jon Breece"
date: "August 28, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.path='figure/')
```

## Project 1: Introduction and Loading of Data
The code below read in the csv of the activity data.
```{r intro, echo=TRUE, warning=FALSE}
activity <- read.csv("activity.csv")
class(activity)
```


## Mean total number of steps taken per day
Below is a table of the total of steps per day and a histogram of the total number of step taken each day.

```{r meanmedian, echo=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
total_steps <- activity %>%
  group_by(date) %>%
  summarise(total = sum(steps))

total_steps

hist(total_steps$total)

mean_steps <- mean(total_steps$total, na.rm = TRUE)

median_steps <- median(total_steps$total, na.rm = TRUE)
```

The mean number of steps is `r as.integer(mean_steps)`, and the median number of steps is `r median_steps`

## Average daily activity pattern
Below is a time series plot of the average number of steps taken per interval.

```{r dailyactivity, echo = TRUE}
library(lattice)
mean_interval <- activity %>%
  group_by(interval) %>%
  summarise(mean = mean(steps, na.rm = TRUE))

xyplot(mean ~ interval, data = mean_interval, type = "l")

arrange(mean_interval,desc(mean))
```

The 5-minute interval with the highest average steps is 835.

## Imputing missing values
The code below substitutes the average number of steps per interval for missing values (e.g., NA).
```{r missingvalues, echo=TRUE, warning=FALSE}

sum(is.na(activity$steps))

activity2 <- left_join(activity,mean_interval,by="interval")
activity2[is.na(activity2$steps),]$steps <- activity2[is.na(activity2$steps),]$mean

total_steps2 <- activity2 %>%
  group_by(date) %>%
  summarise(total = sum(steps))

hist(total_steps2$total)

mean_steps2 <- mean(total_steps2$total)

median_steps2 <- median(total_steps2$total)

```
For the dataset with the imputed values (e.g., activity2), the mean number of steps is `r as.integer(mean_steps2)`, and the median number of steps is `r as.integer(median_steps2)`

## Activity patterns for weekday and weekends

```{r weekdays, echo=TRUE}
activity2$day <- weekdays(as.Date(activity2$date))

weekend <- function(day) {if (day %in% c('Saturday','Sunday')) 
  {
        "weekend"
  } else {
        "weekday"
  }
}

activity2$weekend <- sapply(activity2$day, weekend)
activity2$weekend <- as.factor(activity2$weekend)

mean_interval2 <- activity2 %>%
  group_by(interval, weekend) %>%
  summarise(mean = mean(steps))

xyplot(mean ~ interval | weekend, data= mean_interval2, type = "l")

```

There is a difference in the pattern for weekends and weekdays.